Fast Sorting on Modern Computers

### 2005-08-13 docs/tech_report/heap.tex
changes for testing

It is expected that the number of comparison branches stays the same with
an increased fanout. When the height of the heap halves, the number of
comparisons in each step doubles. However, the proportion of those which are
predictable changes. The key being sifted down the heap is likely to go the
entire way down the tree. Therefore the comparison between the smallest child
and the parent is predictable. However, the comparisons to decide which child is
smallest are not predictable. In a 2-heap, there is one comparison between
children, meaning that half the branches will probably be correctly predicted.
An 8-heap has seven comparisons between children, and will probably have about
3 %TODO does this mean that knuths harmonics applies to this sort of prediction
% thing TODO
mispredictions, as it behaves similarly to selection sort (See \ref{selection
branches logn}). It is possible, therefore, that the 8-heap will have slightly
fewer mispredictions due to the combination of the shallower heap and the larger
fan-out.

@@

\sixplot{base_heapsort_0.eps}{memory_tuned4_heapsort_0.eps}{memory_tuned_heapsort_0.eps}{memory_tuned_heapsort_1.eps}{}{}{base
and memory-tuned heapsorts}
%TESTING TODO
TODO move to testing
In order to explore the behaviour of the individual branches in
heapsort, we performed a second set of branch prediction
experiments. Rather than using SimpleScalar, we instead added
additional statements to the heapsort source code to simulate a
separate bimodal predictor for each branch. This allowed us to measure
each branch individually, whereas SimpleScalar gives only a total
figure for all branches. An additional advantage is that the resulting
compiled simulation is much faster than a full microarchitecture
simulator, and we were thus able to take the average of thousands of
runs (\textbf{TODO find out exact number}), rather than the small
number of runs possible with SimpleScalar. For these experiments,
randomly generated array of size \textbf{xxxxx} were used.

Figure \ref{base and memory-tuned heapsorts-1}(a) shows the behaviour of each
branch in the base heapsort algorithm. The ``creation, has children'' branch
is from line \textbf{xxxxx} of Figure \ref{Simple Base Heapsort}.

%TODO 
\textbf{[David] We need to find some way to describe what each of these
branches is. I don't know myself, so the general reader really, really
won't know.}

>>>>

It is expected that the number of comparison branches stays the same with
an increased fanout. When the height of the heap halves, the number of
comparisons in each step doubles. However, the proportion of those which are
predictable changes. The key being sifted down the heap is likely to go the
entire way down the tree. Therefore the comparison between the smallest child
and the parent is predictable. However, the comparisons to decide which child is
smallest are not predictable. In a 2-heap, there is one comparison between
children, meaning that half the branches will probably be correctly predicted.
An 8-heap has seven comparisons between children, and will probably have about
2.58 mispredictions, as it behaves similarly to selection sort (See \ref{selection
branches harmonics}). It is possible, therefore, that the 8-heap will have slightly
fewer mispredictions (about 36\%) due to the combination of the shallower heap
and the larger fan-out.

@@

\sixplot{base_heapsort_0.eps}{memory_tuned4_heapsort_0.eps}{memory_tuned_heapsort_0.eps}{memory_tuned_heapsort_1.eps}{}{}{base
and memory-tuned heapsorts}

Figure \ref{base and memory-tuned heapsorts-1}(a) shows the behaviour of each
branch in the base heapsort algorithm. The `\cc{Creation, has children}' branch
refers to the check as to whether a key has children, during the creation of the
heap The `\cc{Creation, child comparison}' branch is the comparison between the
two children in the 2-heap, and the `\cc{Creation, child promotion}' refers to
checking if the child is larger than it's parent. The same checks are later made
during the destruction of the heap, and are referred to as `\cc{Destruction, has
children}', `\cc{Destruction, child comparison}' and `\cc{Desruction child
promotion}'.

In the memory-tuned heapsort, the creation phase is simpler, involving merely
one branch, called `\cc{fix-up}' here. The other columns refer to the
destruction phase, where most of the work takes place in memory-tuned heapsoirt.
When a 4-heap is now used, there are three comparisons instead of just one.
Likewise, in an (8-heap), there are seven comparisons.

### 2005-08-13 docs/tech_report/merge.tex
merged in david'd bit to quicksort. added a reference to mergresort. have yet to add replacement policy explanation
changes for testing

\section{Results}
\subsection{Test Parameters}
%TESTING
The level 1 cache is 8k, so 1024 keys from each array can be sorted in the level
1 cache. 262144 keys can be sorted in the level 2 cache. When arrays were
pre-sorted, they were pre-sorted to a size four, or eight if the number of
passes was going to be odd. These were reversed for the double-aligned
multi-mergesort; since it does a k-way merge into the source array, the sorted
cache sized segments need to be in the auxiliary array. An 8-heap was used for
the multi-mergesorts.

@@

It is notable that our prediction of half a miss per key for multi-mergesort,
which was taken from LaMarca, is exactly right. However, our assertion that the
fully associative cache will perform slightly better is false. In actual fact,
the direct mapped cache performs significantly better (see Figure
\vref{multi-mergesort level 2 misses}). This is due to the replacement policy:
our fully associative cache removes the least recently used keys from the cache
to make room, while the direct mapped caches are designed to make best use of
the keys while they are in the cache. They have, effectively, an optimised
replacement policy.

@@

Despite the improvement in cache misses due to multi-mergesort, it still cannot
compensate for the increase in instruction count as a result, and it has a higher
cycle count on the Pentium 4 (see Figure \ref{merge cycles}) as a result. The
other sorts generally perform in $O(NlogN)$ time, with double-aligned tiled mergesort
being the fastest. Note that the alignment of the mergesort may not work as
%TESTING
desired due to the size of the cache. This was optimized for a 2MB cache, while
the Pentium 4 cache is only 256KB. This means that the two arrays are aligned to
the same cache block. However, since the cache is 8-way associative, this is not
serious, and should not result in a lot of thrashing.

>>>>

\section{Results}
\subsection{Test Parameters}
The level 1 cache is 8K, so 1024 keys from each array can be sorted in the level
1 cache. 262144 keys can be sorted in the level 2 cache. When arrays were
pre-sorted, they were pre-sorted to a size four, or eight if the number of
passes was going to be odd. These were reversed for the double-aligned
multi-mergesort; since it does a k-way merge into the source array, the sorted
cache-sized segments need to be in the auxiliary array. An 8-heap was used for
the multi-mergesorts.

@@

\label{mergesort replacement policy}
It is notable that our prediction of half a miss per key for multi-mergesort,
which was taken from LaMarca, is exactly right. However, our assertion that the
fully associative cache will perform slightly better is false. In actual fact,
the direct mapped cache performs significantly better (see Figure
\vref{multi-mergesort level 2 misses}). This is due to the replacement policy:
our fully associative cache removes the least recently used keys from the cache
to make room, while the direct mapped caches are designed to make best use of
the keys while they are in the cache. They have, effectively, an optimised
replacement policy.

@@

Despite the improvement in cache misses due to multi-mergesort, it still cannot
compensate for the increase in instruction count as a result, and it has a higher
cycle count on the Pentium 4 (see Figure \ref{merge cycles}) as a result. The
other sorts generally perform in $O(NlogN)$ time, with double-aligned tiled mergesort
being the fastest. Note that the alignment of the mergesort may not work as
desired due to the size of the cache. This was optimized for a 2MB cache, while
the Pentium 4 cache is only 256KB. This means that the two arrays are aligned to
the same cache block. However, since the cache is 8-way associative, this is not
serious, and should not result in a lot of thrashing.

### 2005-08-13 docs/tech_report/method.tex
changes for testing

%TODO this is going to have to be updated - or is it
\begin{figure}
\scalebox{0.57}{\includegraphics*[0pt,165pt][600pt,802pt]{flowchart}}
\caption{Flowchart describing the development of an algorithm}
\label{Flowchart}
\end{figure}

@@

%TESTING
%TODO a lot of this may be in the first sorts chapter, maybe heapsort.
%discuss: BIG + random.org, 64 bit in lamarca vs 32 bit, do_nothing and how it
%affects cache hits, scaling in the software branch predictors. How to compare
%and contrast varying results (branch vs softwre branhc; P4 vs simplescalar).
%What factors are used for testing: /proc/cpuinfo output, cahce and branch
%predictor data from P4, simplescalar inputs, software predictor, verfication of
%sorts.

Our tests were designed so that we could compare our results to LaMarca's, and
so our results true in the general case. Both our testing method and our data
sets were chosen with these goals in mind.

@@

We chose the a similar approach for our branch predictor caches. We aimed to
simulate bimodal predictors as well as 2 level adaptive, in order to be able to
compare them together. We simulated bimodal predictors with a 512, 1024 and
2048 byte table, and 2 level adaptive predictors with 512, 1024, 2048 and 4096
entries.

@@

\subsection{PaPiex}

Our papiex tests ran significantly faster than either of our other simulations,
and so we were able to run them over a significantly greater quantity of data.
Each sort, and each size from 4096 to 4194304 key, was run 1024 times, using the
system's random number generator to generate data, using seeds of between 0 and
1023.

%TODO did i say 128 byte elsewhere?
These simulations were run a on Pentium 4 1.8Ghz, which had an 8-way, 256K level
2 cache, with a 64-byte cache line. The level 1 caches are seperate 8K 4-way
associative caches.

The instruction pipeline of the Pentium 4 is 20 stages long, and so the branch
misprediction penalty of the Pentium 4 approaches 20 cycles. The Pentium 4's
static branch prediction strategy is backward taken and forward not taken; its
dynamic strategy is unspecified: though \cite{Intel248966-010} mentions a branch
history so it is likely a two-level adaptive predictor.

\subsection{32-bit Integers}
LaMarca used 64-bit \cc{integer}s, as his test machine was a 64-bit system.
Since our machine is 32-bit, and SimpleScalar uses 32-bit addressing on our
architecture, LaMarca's results will not be identical to ours. Rather, the
number of cache misses in his results should be halved to be compared to our;
when this is not the case, it is indicated.

@@

It takes time to fill arrays, and this can distort the relationship between the
results for small sizes, and larger sizes, for which this time is amortized. As
a result, we measured this time, and subtracted it from our results. None of the
graphs show this time.

However, this practice also distorts the results. It removes compulsory cache
misses from the results, so long as the data fits in the cache. When the data
does not fit in the cache, the keys from the start of the array are faulted,
with the effect that capacity misses reoccur, which are not discounted. As a
result, on the left hand side of cache graphs, a number of compulsory misses are
removed, though they are not on the right hand side. Similarly, LaMarca does not
do this, and so this needs to be taken into account when comparing his results
to ours.

@@

As well as the number of cycles, Pentium 4 performance registers can also provide
most of the same information as SimpleScalar, with different parameters. For
example, the cache line of a Pentium 4 machine is 128 bytes whereas 32 byte
cache lines were used in SimpleScalar. The addition of real world data would
strengthen our results.
%TODO not true. we have it

>>>>

\begin{figure}
\scalebox{0.57}{\includegraphics*[0pt,165pt][600pt,802pt]{flowchart}}
\caption{Flowchart describing the development of an algorithm}
\label{Flowchart}
\end{figure}

@@

Our tests were designed so that we could compare our results to LaMarca's, and
so that our results would hold in the general case. Both our testing method and
our data sets were chosen with these goals in mind.

@@

We chose a similar approach for our branch predictor caches. We aimed to
simulate bimodal predictors as well as 2 level adaptive, in order to be able to
compare them together. We simulated bimodal predictors with a 512, 1024 and
2048 byte table, and 2 level adaptive predictors with 512, 1024, 2048 and 4096
entries.

\subsection{Software Predictor}
In order to explore the behaviour of the individual branches, we performed a
second set of branch prediction experiments. Rather than using SimpleScalar, we
instead added additional statements to the source code to simulate a separate
bimodal predictor for each interesting branch. This allowed us to measure each
branch individually, whereas SimpleScalar gives only a total figure for all
branches.

The software predictor results are averaged over ten runs, using arrays
containing 4194304 keys. To compare these results to the SimpleScalar ones, it
is necessary to divide the number of branch mispredictions by 4194304.

\subsection{PaPiex}
Our papiex tests ran significantly faster than either of our other simulations,
and so we were able to run them over a significantly greater quantity of data.
Each sort and each size from 4096 to 4194304 keys was run 1024 times, using the
system's random number generator to generate data, using seeds of between 0 and
1023. The results displayed are averaged over the 1024 runs.

These simulations were run a on Pentium 4 1.6GHz, which had an 8-way, 256K level
2 cache, with a 64-byte cache line. The level 1 caches are 8K 4-way associative
caches, with seperate instruction and data level 1 caches.

The instruction pipeline of the Pentium 4 is 20 stages long, and so the branch
misprediction penalty of the Pentium 4 approaches 20 cycles. The Pentium 4's
static branch prediction strategy is backward taken and forward not taken; its
dynamic strategy is unspecified: though \cite{Intel248966-010} mentions a branch
history so it is likely to be a two-level adaptive predictor.

@@

\subsection{32-bit Integers}
LaMarca used 64-bit \cc{integer}s, as his test machine was a 64-bit system.
Since our machine is 32-bit, and SimpleScalar uses 32-bit addressing on our
architecture, LaMarca's results will not be identical to ours. Rather, the
number of cache misses in his results should be halved to be compared to ours;
when this is not the case, it is indicated.

@@

It takes time to fill arrays, and this can distort the relationship between the
results for small set sizes, and larger set sizes, for which this time is
amortized. As a result, we measured this time, and subtracted it from our
results.

However, this practice also distorts the results. It removes compulsory cache
misses from the results, so long as the data fits in the cache. When the data
does not fit in the cache, the keys from the start of the array are faulted,
with the effect that capacity misses reoccur, which are not discounted. As a
result, on the left hand side of cache graphs, a number of compulsory misses are
removed, though they are not on the right hand side. LaMarca does not do this,
and so this needs to be taken into account when comparing his results to ours.

@@

### 2005-08-13 docs/tech_report/quick.tex
changes for testing

\subsection{More detailed analysis of branch prediction}

@@

Figure \ref{TODO} shows the cycle count of each of the quicksorts,
showing that a pseduo-median-of-seven offers the greatest tradeoff
between the reduced number of branches and instructions, and the
increase in branch mispredictions.

@@

%TODO add a reference to this
\begin{figure}
\includegraphics{plots/quick_median_cycles.eps}
\caption{Empiric cycles count for quicksort with varying medians}
\label{quick_median_cycles}
\end{figure}

>>>>

\subsection{A more detailed look at branch prediction}

@@

Figure \ref{quick_median_cycles} shows the cycle count of each of the quicksorts,
showing that a pseduo-median-of-seven offers the greatest tradeoff
between the reduced number of branches and instructions, and the
increase in branch mispredictions.

@@

\begin{figure}
\includegraphics{plots/quick_median_cycles.eps}
\caption{Empiric cycles count for quicksort with varying medians}
\label{quick_median_cycles}
\end{figure}
